---
title: "BCB420 Assignment1"
author: "Jiayan Wang"
output:
  html_document:
    df_print: paged
---

## 1. Introduction

Lung cancer is the one of the leading cause of death from cancer and non small cell lung cancer(NSCLC) is the most common type of lung cancer, account for 85% of all lung cancer.Morever, statistics shows that the overall 5-year survival rate for NSCLC is under 20%. Recent studies have revealed the important role of G9a, a histone methyltransferase, in various cancer cells. G9a is responsible for the mono- and di-methylation of H3K9 and its overexpression was ovserved in most lung cancer cells including

Therefore, I choose the dataset that investigate the role of dysregulated histone methyltransferase G9a in two human non-small cell lung cancer (NSCLC) cell lines. The mechanistic investigation of G9A knockdown and its role in in anti-tumor activity might be able to uncover the possible therapeutic target in NSCLC diagnosis and treatment.




## 2. Get GSE object and platform information:

 * R packages that required for data processing
 
```{r, message=FALSE}
library(BiocManager)
library(GEOquery)
library(edgeR)
library(biomaRt)
library(Biobase)
```
 
 * Download and parse GSE object from NCBI GEO

```{r Get GSE object, message=FALSE}
gse <- getGEO("GSE113493",GSEMatrix=FALSE)
current_gpl <- names(GPLList(gse))[1]
current_gpl_info <- Meta(getGEO(current_gpl))
```
* My dataset: GSE113493

* Title: `r gse@header$title`

* Platform: `r gse@gpls$GPL11154@header$geo_accession`

* Platform title: `r gse@gpls$GPL11154@header$title`

* Organism: `r gse@gpls$GPL11154@header$organism`

* Experiment type: `r gse@gpls$GPL11154@header$technology`

* Number of GEO datasets that use this techology: `r length(current_gpl_info$series_id)`
 
* Number of GEO samples that use this technology: `r length(current_gpl_info$sample_id)`
 


* Overall Design: Two non-small cell lung cancer(NSCLC) cell lines (A549 and H1299) were first treated with G9A inhibitor to silence G9a expression. Then cells were transfected into two condition: one with control siRNA and other with G9a specific siRNA. Then differential gene expression of cells upon knockdown of G9a were performed to see if these pathways contains G9a-mediated cancer expression.  
 
## 3. Get the expression data

```{r}
sfiles = getGEOSuppFiles('GSE113493')
fnames = rownames(sfiles)
exp_data = read.delim(fnames[1],header=TRUE,
                      check.names = FALSE)
colnames(exp_data)[1] = "gname"
head(exp_data)
```


* __create a table to show data characteristics in the sampling__
```{r}
samples <- data.frame(lapply(colnames(exp_data)[2:9],
                             FUN=function(x){
                               x <- gsub("(-[G])", "_G9a+-", x)
                               x <- gsub("(-[C])", "_Control-", x)
                               unlist(strsplit(x, split = "_"))[c(2,3)]}))
colnames(samples) <- colnames(exp_data)[2:9]
rownames(samples) <- c("cell","control_vs_treatment")
samples <- data.frame(t(samples))
head(samples)
```

The dataset contains `r nrow(exp_data)` genes of interest from `r nrow(samples)` samples of `r length(unique(samples$cell))` NSCLC cell lines.



## 4. clean the data
* __get number of rows(genes) in the dataset__
```{r}
dim(exp_data)
```
There are `r nrow(exp_data)` genes measured from 8 experiments: 2 control and 2 treated with Ga9 for 2 cell lines. (2*2*2 = 8)

* __get frequence counts for each gene__
```{r}
gene_counts <- sort(table(exp_data$gname),decreasing = TRUE)
head(gene_counts)
```
There are no duplicates in the data, that is, expression value is unique for each gene.

* __filter weakly expressed features__
```{r message=FALSE}
library(edgeR)
cpms <- cpm(y =exp_data[,2:9])
rownames(cpms) <- exp_data[,1]
keep = rowSums(cpms >1) >=3
exp_data_filtered = exp_data[keep,]
colnames(exp_data_filtered)[1] = "gname"
dim(exp_data) #before filtering
dim(exp_data_filtered) #after filtering
```
`r nrow(exp_data)- nrow(exp_data_filtered)` genes with low-count(less than 3) were filtered out. This helps removes noises and will enrich for true differential expression.

## 5. data distribution


* __visualize our data__
```{r warning=FALSE}
data2plot <- log2(cpm(exp_data_filtered[,2:9]))
bp <- boxplot(data2plot, xlab = "Samples", ylab = "log2 CPM",
        las = 2, cex = 0.5, cex.lab = 0.5,
        cex.axis = 0.5, main = "Ga9 RNASeq Samples")

#draw the median on each box plot
abline(h = median(apply(data2plot, 2, median)), col = "green", lwd = 2, lty = "dashed")

```

The boxplot shows that the median of each sample are very close to each other. This is consistant with the fact that the dta has been normalized already. 
There are also a few outliers in the boxplot and some of them are too extreme to show, so we can have a look of them.
* There are `r length(bp$out)` outliers and most of them are due to low counts(less than 2) in the raw count dataset
* `r sum(bp$out == "-Inf")` ouliers shows "-Inf" which means negative infinity. This is due to 0 counts in the dataset.
* Outliers were not removed at this stage. I



* __density plot__
```{r}
counts_density <- apply(log2(cpm(exp_data_filtered[,2:9])), 2, density)
  #calculate the limits across all the samples
    xlim <- 0; ylim <- 0
    for (i in 1:length(counts_density)) {
      xlim <- range(c(xlim, counts_density[[i]]$x));
      ylim <- range(c(ylim, counts_density[[i]]$y))
    }
    cols <- rainbow(length(counts_density))
    ltys <- rep(1, length(counts_density))
    #plot the first density plot to initialize the plot
    plot(counts_density[[1]], xlim=xlim, ylim=ylim, type="n",
         ylab="Smoothing density of log2-CPM", main="", cex.lab = 0.85)
    #plot each line
    for (i in 1:length(counts_density)) lines(counts_density[[i]], col=cols[i], lty=ltys[i])
    #create legend
    legend("topright", colnames(data2plot),  
           col=cols, lty=ltys, cex=0.75,
           border ="blue",  text.col = "green4",
           merge = TRUE, bg = "gray90")
```

## 6. Data Normalization

The main aim for data normaliztion is to make the expression value comparable between different samples. Trimmed Mean of M-values” normalization (TMM) adjustment is considered appropriate as it use one sample as a reference and estimate correction factors for each samples.(ref:https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0206312) TMM method is implemented using edgeR package.

The original data has been normalized. According to the paper, the dataset has been normalized using Illumina Genome Analyzer II.But I still applied TMM to see if there is any difference after 

Therefore, no significant difference observed after applying TMM to the dataset.  


```{r}
filtered_data_matrix <- as.matrix(exp_data_filtered[,2:9])
rownames(filtered_data_matrix) <- exp_data_filtered$gname
d = DGEList(counts=filtered_data_matrix, group=samples$cell_type)
#get the normalized data
normalized_counts <- cpm(d)
```

* __desity plot after normalization__
```{r}
normalized_counts_density <- apply(log2(normalized_counts[,2:8]), 2, density)

  #calculate the limits across all the samples
    xlim <- 0; ylim <- 0
    for (i in 1:length(normalized_counts_density)) {
      xlim <- range(c(xlim, normalized_counts_density[[i]]$x));
      ylim <- range(c(ylim, normalized_counts_density[[i]]$y))
    }
    cols <- rainbow(length(normalized_counts_density))
    ltys <- rep(1, length(normalized_counts_density))
    #plot the first density plot to initialize the plot
    plot(normalized_counts_density[[1]], xlim=xlim, ylim=ylim, type="n",
         ylab="Smoothing density of log2-CPM", main="Density Plot after Normalization", cex.lab = 0.85)
    #plot each line
    for (i in 1:length(normalized_counts_density)) lines(normalized_counts_density[[i]], col=cols[i], lty=ltys[i])
    #create legend
    legend("topright", colnames(data2plot),  
           col=cols, lty=ltys, cex=0.75,
           border ="blue",  text.col = "green4",
           merge = TRUE, bg = "gray90")
```

* __MDS plot to show the distribution of samples__
```{r}
plotMDS(d, labels=rownames(samples),
  col = c("darkgreen","blue")[factor(samples$control_vs_treatment)],cex = 0.8,xlim= range(-6,6),
  main="Multidimensional Scaling Plot")
```

The MDS plot show that samples have smaller distance within same cell line within both control and G9a group. The overlapping samples are replicates that showing highes similarities. 




## 7. Identifier mapping

The dataset has been already mapped with HGNC identifiers. They were further validated through HGNChelper package.

```{r, message=FALSE}
if (!require("HGNChelper")) install.packages("HGNChelper")
library(HGNChelper)
```

```{r, warning=FALSE}
hgnc_check <- HGNChelper::checkGeneSymbols(exp_data_filtered$gname,species = "human")
unknown <- hgnc_check[hgnc_check$Approved == FALSE & is.na(hgnc_check$Suggested.Symbol),]
incorrect<- hgnc_check[hgnc_check$Approved == FALSE & !is.na(hgnc_check$Suggested.Symbol),]
head(unknown)
head(incorrect)
```


* `r nrow(unknown)` identifiers were shown unmatched according to HGNC gene name checker. So they were removed from the dataset.

* `r nrow(incorrect)` identifiers were shown incorrect and were replaced by suggested symbols. 

* For identifier mapping, we have `r (nrow(exp_data_filtered)-nrow(unknown))/nrow(exp_data_filtered)*100` coverage of the dataset.

* Overall, we have `r (nrow(exp_data_filtered)-nrow(unknown))/ nrow(exp_data)*100` coverage of the dataset at this stage.


## 8. Reference

[1]GEO Accession viewer. (n.d.). Retrieved February 4, 2020, from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE113493

[2]Lung Cancer - Non-Small Cell: Statistics | Cancer.Net. (n.d.). Retrieved February 4, 2020, from https://www.cancer.net/cancer-types/lung-cancer-non-small-cell/statistics

[3]Zhang, K., Wang, J., Yang, L., Yuan, Y.-C., Tong, T. R., Wu, J., … Raz, D. J. (2018). Targeting histone methyltransferase G9a inhibits growth and Wnt signaling pathway by epigenetically regulating HP1α and APC2 gene expression in non-small cell lung cancer. Molecular Cancer, 17(1), 153. https://doi.org/10.1186/s12943-018-0896-8

[4]Tachibana, M., Sugimoto, K., Nozaki, M., Ueda, J., Ohta, T., Ohki, M., … Shinkai, Y. (2002). G9a histone methyltransferase plays a dominant role in euchromatic histone H3 lysine 9 methylation and is essential for early embryogenesis. Genes and Development, 16(14), 1779–1791. https://doi.org/10.1101/gad.989402
